Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-03 paper/paper.tex
More edits and one giant comment

We next take a closer look at the remaining rules for the inductive relation
$[[Interp I i A S]]$ in Figure~\ref{fig:logrel}.
\Rref{I-Void, I-Bool} capture terms that \emph{behave} like the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type only contains terms that
reduce to $[[true]]$ or $[[false]]$. Note that the characterization
of $[[Bool]]$ (and other inhabited types) in our logical relation does not always correspond to
well-typed or even closed terms. For example, the term $[[if false (#
Void true #)
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation when we extend our logical
relation to prove the existence of $\beta$-normal forms (Section~\ref{sec:extension}).
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresponds to the
set of terms that reduce to $[[refl]]$ when $[[a <=> b]]$ also holds and
otherwise corresponds to the empty set. Conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce types in order
to assign meanings. Recall the type expression $[[(\ x . x )
Bool]]$. \Rref{I-Red} says that to know that
$[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\ x . x) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ x . x) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complex rule in our logical
relation. Instead of explaining it directly, we first consider the following
simplified version, \rref{I-PiAlt}, that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\Rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
the rule says that
the function type $[[Pi x : A . B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation $[[S0]]$. In its conclusion, the 
interpretation of $[[Pi x : A . B]]$ is the set of terms $[[b]]$, such that
for all $[[a in S]]$, where $[[S]]$ is an interpretation of $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a / x}]]$ where $[[a in S]]$).

>>>>

We next take a closer look at the remaining rules for the inductive relation
$[[Interp I i A S]]$ in Figure~\ref{fig:logrel}.
\Rref{I-Void, I-Bool} capture terms that \emph{behave} like the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. In particular,
the $[[Void]]$ type has no inhabitants, while the $[[Bool]]$ type only contains
terms that reduce to $[[true]]$ or $[[false]]$. Note that the characterization
of $[[Bool]]$ (and other inhabited types) in our logical relation does not
always correspond to well-typed or even closed terms. For example, the term
$[[if false (# Void true #) true]]$ is ill typed under the empty context,
but still belongs to the set $[[{ a | a =>+ true \/ a =>+ false }]]$ since it
evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation when we extend our logical
relation to prove the existence of $\beta$-normal forms in Section~\ref{sec:extension}.
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}
\jc{Should there be an explanation of what Kripke-style logical relations are here?}

\Rref{I-Eq} states that an equality type $[[a ~ b : A]]$ corresponds to the
set of terms that reduce to $[[refl]]$ when $[[a <=> b]]$ also holds and
otherwise corresponds to the empty set. Conditions like $[[a <=> b]]$ are
typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce types in order
to assign meanings. Recalling expression $[[(\ x . x ) Bool]]$, \rref{I-Red}
says that to know that $[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$,
it suffices to show that $[[Interp I i Bool S]]$, since
$[[(\ x . x) Bool => Bool]]$.
The derivation for $[[Interp I i (\ x . x) Bool { a | a =>+ true \/ a =>+ false }]]$
therefore follows by composing \rref{I-Red} and \rref{I-Bool}.

\Rref{I-Pi} is the most complex rule in our logical relation.
Consider first the following simplified \rref{I-PiAlt}, which follows from \rref{I-Pi}.

\begin{center}
  \drule[]{I-PiAlt}
\end{center}

\Rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The preconditions of
the rule says that the function type $[[Pi x : A . B]]$ has an interpretation
if its input type $[[A]]$ can be interpreted as some set $[[S]]$, and for all
terms $[[a in S]]$, the type $[[B {a / x}]]$, obtained by substituting $[[a]]$
into the output type $[[B]]$, has some semantic interpretation $[[S0]]$. In its
conclusion, the  interpretation of $[[Pi x : A . B]]$ is the set of terms
$[[b]]$ such that for all $[[a in S]]$, where $[[S]]$ is an interpretation of
$[[A]]$, the application form $[[b a]]$ belongs to all possible interpretations
of $[[B {a  / x}]]$. The second precondition ensures at least one interpretation
exists for each $[[B {a / x}]]$ where $[[a in S]]$.

\jc{I think at this point it's unclear why \rref{I-PiAlt} is shaped the way it is.
I found myself asking: If in the second precondition we've established the existence of $[[S0]]$,
why do we need to quantify over all interpretations $[[S0]]$?
In fact, instead of the second precondition,
why can't split it into two and say that there exists some family of sets
$S_0 : \forall a, a \in S \to [[PowerSet STm]]$
such that $\forall a, (m : a \in S) \to [[Interp I i B { a / x } (# S0 a m #)]]$,
then conclude that the intepretation of the function type yields a set over $[[b]]$
such that $\forall a, (m : a \in S) \to [[b in S0 a m]]$?
As far as I can tell, the interpretation would be well formed,
but proving the fundamental theorem would then require something like indefinite description,
since the induction hypothesis would yield $\forall a, a \in S \to \exists S_0, \dots$,
the mere existence of such a family of sets,
while the constructor requires the actual family of sets itself.
This is subtle but maybe warrants explaining.}
